A tool set for random number generation on GPUs in R

### 2021-11-30 pkg/gpuRandom/inst/documents/paper1_2021_5/paper1.Rnw
update paper1
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/diseasemapping@3011 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

 <<monthdata,eval=TRUE,echo=FALSE, cache=TRUE, fig.pos='h'>>=
library(knitr)
knitr::kable(month,format="latex", align = c("rrrrrrrrrrrr"), 
  caption = "Monthly birth anomaly data\\label{tab:month}", booktabs=TRUE,linesep = "") %>%
  kable_styling(full_width = F, position = "center")#latex_options = "HOLD_position")
@

@@

\subsection{A summary of the results}
We summarized the comparison results in the table \ref{tab:summary} and plot the test statistics in Figure \ref{fig4}.
<<summarycompare, echo=FALSE, eval=TRUE, cache=TRUE, fig.pos='h'>>=
library(kableExtra)
dt <- data.frame(
  var1 = c("B", "1M", "10M", "1M", "10M"),
  var2 = c('Intel 2.5ghz', 0.403804, 0.0001251, 10.74,  63.24),
  var3 = c('AMD Radeon', 0.403507, 0.0001274,   2.28,    10.82),
  var4 = c('Intel 3.7ghz', 0.4035606,0.0001202, 15.00,  91.58),
  var5 = c('NVIDIA V100', 0.403507,0.0001274,   0.72,   4.18),
  var6 = c('Data','month','week','month','week')
)

@@

<<swissRainBoundary, echo=TRUE>>=
data("swissRain", package="geostatsp")
myRaster = geostatsp::squareRaster(swissBorder, 80)
myRaster
@
Step 2, create 5 parameter sets as a small example, in practical studies, there can be hundreds or thousands of parameter sets. Then, convert the matrix of parameters to a ``vclMatrix'' for later use by \fct{gpuBatchMatrix::maternGpuParam}.
<<setup paramsBatch, eval=TRUE,echo=FALSE>>=
params = 
rbind(c(shape=1.25, range=50*1000, variance = 1.5, nugget = 0,anisoRatio = 1, anisoAngleRadians = 0), 
c(shape=2.15, range=60*1000, variance = 2, nugget = 0, anisoRatio = 4, anisoAngleRadians = pi/7),
c(shape=0.6, range=30*1000, variance = 2, nugget = 0, anisoRatio = 2, anisoAngleRadians = pi/7),
c(shape=3, range=30*1000, variance = 2, nugget = 0, anisoRatio = 2, anisoAngleRadians = pi/7)
)
#c(shape=2.15, range=40*1000, variance = 2, nugget = 0, anisoRatio = 2, anisoAngleRadians = pi/4))
@
<<show parameterBatch,eval=TRUE>>=
params
#paramsGpu = gpuBatchMatrix::maternGpuParam(myParamsBatch, type=theType)
@
Step 3, compute Mat\'ern covariance matrices using \fct{gpuBatchMatrix::maternBatch}, the returned Mat\'ern covariance matrices %$\Sigma_1, \Sigma_2, \dots, \Sigma_5$ correspond to the 5 parameter sets respectively, the matrices 
are each of size $4800 \times 4800$ and are stacked by row in the output \code{maternCov}. 
<<r outputBatchCreate, eval=TRUE>>=
#maternCov = vclMatrix(0, nrow(paramsGpu)*nrow(coordsGpu), 
#                         nrow(coordsGpu),type=theType)
maternCov = gpuBatchMatrix::maternBatch(
  params, myRaster,          
  Nglobal=c(128,64), Nlocal=c(16,4))
dim(maternCov)
@
Step 4, the first argument \code{A} in \fct{gpuBatchMatrix::cholBatch} specifies the object to take Cholesky decomposition. Computed unit lower triangular matrices $L_i$'s are stacked by row and stored in \code{maternCov}. Each row of \code{diagMat} stores the diagonal values of each $D_i$, for example, if each batch $\Sigma_i$ is of size $n \times n$, then each batch $L_i$ is $n \times n$, and each batch $D_i$ is $1 \times n$.
<<r cholBach, eval=TRUE>>=
diagMat = gpuBatchMatrix::cholBatch(maternCov, 
                          Nglobal= c(128, 8), Nlocal= c(32, 8))
@
\begin{gather}
\begin{bmatrix} \Sigma_{1} \\ \Sigma_2 \\ \Sigma_3 \\ \vdots
\end{bmatrix}
 \rightarrow
 \begin{bmatrix}
  L_{1} \\ L_{2} \\L_{3} \\ \vdots
  \end{bmatrix} \text{and}
  \begin{bmatrix}
  D_{1} \\ D_{2} \\D_{3} \\ \vdots
  \end{bmatrix}
\end{gather}
Step 5, generate 2 standard Gaussian random vectors \code{zmatGpu}$=(Z_1, Z_2)$ using \fct{clrng::rnorm}, in which \code{c(nrow(materCov),2)} specifies the number of rows and columns of \code{zmatGpu}.
<<r randomNormalsSim1, eval=TRUE, cache=FALSE>>=
streamsGpu <- createStreamsGpu(n=128*64)
@
<<r randomNormalsSim2, eval=TRUE, cache=FALSE>>=
zmatGpu = clrng::rnorm(
  c(nrow(maternCov),2), streams=streamsGpu, 
                Nglobal=c(128,2),
                 type = theType)
@
Step 6, use \fct{gpuBatchMatrix::multiplyLowerDiagonalBatch} to compute $U = L * D^{(1/2)}* Z$ in batches, see the following illustration. \code{simMat} is the output matrix for $U$, \code{maternCov}, \code{diagMat}, and \code{zmatGpu} correspond to the matrices of $L$, $D$ and $Z=(Z_1, Z_2)$ respectively.
<<r multLowerdiag_LDZ, eval=TRUE, cache=FALSE, results='hide'>>=
simMat = vclMatrix(0, nrow(zmatGpu), ncol(zmatGpu), 
                   type = gpuR::typeof(zmatGpu))

>>>>

 <<monthdata,eval=TRUE,echo=FALSE, cache=TRUE, fig.pos='h'>>=
library(knitr)
knitr::kable(month,format="latex", align = c("rrrrrrrrrrrr"), 
  caption = "\\label{tab:month}Monthly birth anomaly data", booktabs=TRUE,linesep = "") %>%
  kable_styling(full_width = F, position = "center")#latex_options = "HOLD_position")
@

@@

\subsection{A summary of the results}
We summarized the comparison results in the table \ref{tab:summary} and plot the test statistics in Figure %\ref{fig4}.
<<summarycompare, echo=FALSE, eval=TRUE, cache=TRUE, fig.pos='h'>>=
library(kableExtra)
dt <- data.frame(
  var1 = c("B", "1M", "10M", "1M", "10M"),
  var2 = c('Intel 2.5ghz', 0.403804, 0.0001251, 10.74,  63.24),
  var3 = c('AMD Radeon', 0.403507, 0.0001274,   2.28,    10.82),
  var4 = c('Intel 3.7ghz', 0.4035606,0.0001202, 15.00,  91.58),
  var5 = c('NVIDIA V100', 0.403507,0.0001274,   0.72,   4.18),
  var6 = c('Data','month','week','month','week')
)

@@

<<swissRainBoundary, echo=TRUE>>=
data("swissRain", package="geostatsp")
myRaster = geostatsp::squareRaster(swissBorder, 80)
myRaster
@
Step 2, create 5 parameter sets as a small example, in practical studies, there can be hundreds or thousands of parameter sets. Then, convert the matrix of parameters to a ``vclMatrix'' for later use by \fct{gpuBatchMatrix::maternGpuParam}.
<<setup paramsBatch, eval=TRUE,echo=FALSE>>=
params = 
rbind(c(shape=1.25, range=50*1000, variance = 1.5, nugget = 0,anisoRatio = 1, anisoAngleRadians = 0), 
c(shape=2.15, range=60*1000, variance = 2, nugget = 0, anisoRatio = 4, anisoAngleRadians = pi/7),
c(shape=0.6, range=30*1000, variance = 2, nugget = 0, anisoRatio = 2, anisoAngleRadians = pi/7),
c(shape=3, range=30*1000, variance = 2, nugget = 0, anisoRatio = 2, anisoAngleRadians = pi/7)
)
#c(shape=2.15, range=40*1000, variance = 2, nugget = 0, anisoRatio = 2, anisoAngleRadians = pi/4))
@
<<show parameterBatch,eval=TRUE>>=
params
#paramsGpu = gpuBatchMatrix::maternGpuParam(myParamsBatch, type=theType)
@
Step 3, compute Mat\'ern covariance matrices using \fct{gpuBatchMatrix::maternBatch}, the returned Mat\'ern covariance matrices %$\Sigma_1, \Sigma_2, \dots, \Sigma_5$ correspond to the 5 parameter sets respectively, the matrices 
are each of size $4800 \times 4800$ and are stacked by row in the output \code{maternCov}. 
<<r outputBatchCreate, eval=TRUE>>=
#maternCov = vclMatrix(0, nrow(paramsGpu)*nrow(coordsGpu), 
#                         nrow(coordsGpu),type=theType)
maternCov = gpuBatchMatrix::maternBatch(
  params, myRaster,          
  Nglobal=c(128,64), Nlocal=c(16,4))
dim(maternCov)
@
Step 4, the first argument \code{A} in \fct{gpuBatchMatrix::cholBatch} specifies the object to take Cholesky decomposition. Computed unit lower triangular matrices $L_i$'s are stacked by row and stored in \code{maternCov}. Each row of \code{diagMat} stores the diagonal values of each $D_i$, for example, if each batch $\Sigma_i$ is of size $n \times n$, then each batch $L_i$ is $n \times n$, and each batch $D_i$ is $1 \times n$.
<<r cholBach, eval=TRUE>>=
diagMat = gpuBatchMatrix::cholBatch(maternCov, 
                          Nglobal= c(128, 8), Nlocal= c(32, 8))
@
\begin{gather}
\begin{bmatrix} \Sigma_{1} \\ \Sigma_2 \\ \Sigma_3 \\ \vdots
\end{bmatrix}
 \rightarrow
 \begin{bmatrix}
  L_{1} \\ L_{2} \\L_{3} \\ \vdots
  \end{bmatrix} \text{and}
  \begin{bmatrix}
  D_{1} \\ D_{2} \\D_{3} \\ \vdots
  \end{bmatrix}
\end{gather}
Step 5, generate 2 standard Gaussian random vectors \code{zmatGpu}$=(Z_1, Z_2)$ using \fct{clrng::rnorm}, in which \code{c(nrow(materCov),2)} specifies the number of rows and columns of \code{zmatGpu}.
<<r randomNormalsSim1, eval=TRUE, cache=FALSE>>=
streamsGpu <- createStreamsGpu(n=128*64)
@
<<r randomNormalsSim2, eval=TRUE, cache=FALSE>>=
zmatGpu = clrng::rnorm(
  c(nrow(maternCov),2), streams=streamsGpu, 
                Nglobal=c(128,64),
                type = theType)
@
Step 6, use \fct{gpuBatchMatrix::multiplyLowerDiagonalBatch} to compute $U = L * D^{(1/2)}* Z$ in batches, see the following illustration. \code{simMat} is the output matrix for $U$, \code{maternCov}, \code{diagMat}, and \code{zmatGpu} correspond to the matrices of $L$, $D$ and $Z=(Z_1, Z_2)$ respectively.
<<r multLowerdiag_LDZ, eval=TRUE, cache=FALSE, results='hide'>>=
simMat = vclMatrix(0, nrow(zmatGpu), ncol(zmatGpu), 
                   type = gpuR::typeof(zmatGpu))

### 2021-11-30 pkg/gpuRandom/inst/documents/paper1_2021_5/paper1.tex
update paper1
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/diseasemapping@3011 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

%\usepackage{float}
%\usepackage{framed}
\usepackage{subcaption}
\renewcommand{\subfloat}[2][need a sub-caption]{ \subcaptionbox{#1}{#2} }
\usepackage{amsmath}
\usepackage[ruled,vlined]{algorithm2e}
\usepackage{listings}
\usepackage[shortcuts]{extdash}
\usepackage{booktabs}

@@

We generate a large-size matrix of 100 million double-precision Gaussian random numbers, and compare the run-time between using \fct{stats::rnorm} and \fct{clrng::rnorm}. The best performance we have seen using \fct{clrng::rnorm} is around 120 times faster with $512 \times 128$ work-items than using the \fct{stats::rnorm}. The difference in elapsed time becomes larger when matrix size goes larger.
\begin{CodeChunk}
\begin{CodeOutput}
##    user  system elapsed 
##   0.077   0.000   0.076
\end{CodeOutput} 
\end{CodeChunk}

\begin{CodeChunk}
\begin{CodeOutput}
##    user  system elapsed 
##   7.103   0.651   7.750
\end{CodeOutput} 
\end{CodeChunk}

@@

Below is an example that produces a matrix of Exponential random numbers with expectation equal to 1.
\begin{CodeChunk}
\begin{CodeOutput}
##       [,1]  [,2]  [,3]  [,4]
## [1,] 0.343 0.364 0.542 0.054
## [2,] 1.113 3.642 0.200 2.770
\end{CodeOutput} 
\end{CodeChunk}

@@

\begin{table}

\caption{\label{tab:monthdata}Monthly birth anomaly data\label{tab:month}}
\centering
\begin{tabular}[t]{lrrrrrrrrrrrr}
\toprule
  & Ane & Men & Cya & Her & Omp & Gas & Lim & Cle & Pal & Dow & Chr & Hyp\\
\midrule
Jan & 29 & 55 & 172 & 46 & 39 & 73 & 48 & 183 & 77 & 103 & 102 & 174\\
Feb & 25 & 45 & 175 & 35 & 31 & 55 & 34 & 142 & 81 & 115 & 100 & 180\\
Mar & 31 & 48 & 182 & 41 & 47 & 72 & 40 & 200 & 86 & 90 & 96 & 180\\
Apr & 34 & 45 & 186 & 36 & 32 & 75 & 42 & 173 & 56 & 87 & 90 & 193\\
May & 33 & 40 & 187 & 46 & 24 & 80 & 35 & 180 & 75 & 91 & 100 & 197\\
Jun & 34 & 48 & 189 & 35 & 33 & 75 & 45 & 154 & 74 & 102 & 100 & 182\\
Jul & 26 & 43 & 198 & 34 & 21 & 74 & 36 & 179 & 79 & 86 & 92 & 193\\
Aug & 24 & 41 & 189 & 44 & 43 & 62 & 48 & 183 & 88 & 109 & 94 & 194\\
Sept & 34 & 44 & 147 & 40 & 37 & 66 & 36 & 158 & 73 & 112 & 103 & 196\\
Oct & 25 & 43 & 207 & 45 & 31 & 65 & 49 & 181 & 77 & 108 & 115 & 220\\
Nov & 36 & 55 & 188 & 39 & 39 & 62 & 43 & 144 & 68 & 98 & 79 & 173\\
Dec & 23 & 48 & 196 & 31 & 31 & 71 & 31 & 177 & 86 & 86 & 73 & 156\\
\bottomrule
\end{tabular}
\end{table}

\begin{CodeChunk}
\begin{CodeOutput}
##    user  system elapsed 
##   0.316   0.000   0.316
\end{CodeOutput}
\begin{CodeOutput}
## [1] -47955
\end{CodeOutput}
\begin{CodeOutput}
## [1] 1015808
\end{CodeOutput}
\begin{CodeOutput}
## [1] 410825
\end{CodeOutput}
\begin{CodeOutput}
## [1] 0.404
\end{CodeOutput} 
\end{CodeChunk}

@@

\begin{CodeChunk}
\begin{CodeOutput}
##    user  system elapsed 
##   0.137   0.000   0.137
\end{CodeOutput}
\begin{CodeOutput}
## [1] 0.4
\end{CodeOutput} 
\end{CodeChunk}

@@

\begin{table}

\caption{\label{tab:weekdata}Day-of-week birth anomaly data\label{tab:week}}
\centering
\begin{tabular}[t]{lrrrrrrrrrrrr}
\toprule
  & Ane & Men & Cya & Her & Omp & Gas & Lim & Cle & Pal & Dow & Chr & Hyp\\
\midrule
Mon & 30 & 34 & 173 & 37 & 23 & 80 & 49 & 191 & 83 & 122 & 109 & 216\\
Tue & 60 & 121 & 383 & 80 & 83 & 131 & 71 & 349 & 146 & 164 & 168 & 352\\
Wed & 51 & 106 & 417 & 92 & 73 & 145 & 72 & 333 & 136 & 179 & 196 & 351\\
Thu & 60 & 86 & 362 & 69 & 74 & 120 & 85 & 326 & 132 & 220 & 187 & 359\\
Fri & 52 & 94 & 347 & 87 & 59 & 123 & 68 & 323 & 145 & 170 & 166 & 345\\
Sat & 52 & 63 & 323 & 67 & 64 & 135 & 73 & 316 & 170 & 189 & 188 & 357\\
Sun & 49 & 51 & 211 & 40 & 32 & 96 & 69 & 216 & 108 & 143 & 130 & 258\\
\bottomrule
\end{tabular}
\end{table}

\begin{CodeChunk}
\begin{CodeOutput}
##    user  system elapsed 
##   3.854   0.011   3.864
\end{CodeOutput}
\begin{CodeOutput}
## [1] -54990
\end{CodeOutput}
\begin{CodeOutput}
## [1] 1e+07
\end{CodeOutput}
\begin{CodeOutput}
## [1] 1193
\end{CodeOutput}
\begin{CodeOutput}
## [1] 0.000119
\end{CodeOutput} 
\end{CodeChunk}

@@

\begin{CodeChunk}
\begin{CodeOutput}
##    user  system elapsed 
##   0.009   0.000   0.009
\end{CodeOutput}
\begin{CodeOutput}
## [1] 0.000999
\end{CodeOutput} 
\end{CodeChunk}

\subsection{A summary of the results}
We summarized the comparison results in the table \ref{tab:summary} and plot the test statistics in Figure \ref{fig4}.
\begin{table}

@@

\begin{CodeChunk}
\begin{CodeInput}
R> data("swissRain", package = "geostatsp")
R> myRaster = geostatsp::squareRaster(swissBorder, 60)
R> myRaster
\end{CodeInput}
\begin{CodeOutput}
## class      : RasterLayer 
## dimensions : 38, 60, 2280  (nrow, ncol, ncell)
## resolution : 5811, 5811  (x, y)
## extent     : 2485351, 2833996, 1075013, 1295822  (xmin, xmax, ymin, ymax)
## crs        : +proj=somerc +lat_0=46.9524055555556 +lon_0=7.43958333333333 +k_0=1 +x_0=2600000 +y_0=1200000 +ellps=GRS80 +units=m +no_defs
\end{CodeOutput} 
\end{CodeChunk}

@@

Step 3, compute Mat\'ern covariance matrices using \fct{gpuBatchMatrix::maternBatch}, the returned Mat\'ern covariance matrices %$\Sigma_1, \Sigma_2, \dots, \Sigma_5$ correspond to the 5 parameter sets respectively, the matrices 
are each of size $4800 \times 4800$ and are stacked by row in the output \code{maternCov}. 
\begin{CodeChunk}
\begin{CodeOutput}
## [1] 9120 2280
\end{CodeOutput} 
\end{CodeChunk}

>>>>

%\usepackage{float}
%\usepackage{framed}
\usepackage{subcaption}
\renewcommand{\subfloat}[2][need a sub-caption]{ \subcaptionbox{#1}{#2} }
\usepackage{amsmath}
\usepackage[ruled,vlined]{algorithm2e}
\usepackage{listings}
\usepackage[shortcuts]{extdash}
\usepackage{booktabs}
\usepackage[dvipsnames]{xcolor}
\usepackage{pdfpages}

@@

We generate a large-size matrix of 100 million double-precision Gaussian random numbers, and compare the run-time between using \fct{stats::rnorm} and \fct{clrng::rnorm}. The best performance we have seen using \fct{clrng::rnorm} is around 120 times faster with $512 \times 128$ work-items than using the \fct{stats::rnorm}. The difference in elapsed time becomes larger when matrix size goes larger.
\begin{CodeChunk}
\begin{CodeOutput}
##    user  system elapsed 
##   0.060   0.009   0.069
\end{CodeOutput} 
\end{CodeChunk}

\begin{CodeChunk}
\begin{CodeOutput}
##    user  system elapsed 
##    6.61    0.80    7.41
\end{CodeOutput} 
\end{CodeChunk}

@@

Below is an example that produces a matrix of Exponential random numbers with expectation equal to 1.
\begin{CodeChunk}
\begin{CodeOutput}
##       [,1]  [,2]  [,3]  [,4]
## [1,] 0.095 0.132 0.679 0.458
## [2,] 0.275 0.132 1.118 0.111
\end{CodeOutput} 
\end{CodeChunk}

@@

{\ttfamily\noindent\bfseries\color{errorcolor}{\#\# Error in knitr::kable(month, format = "{}latex"{}, align = c("{}rrrrrrrrrrrr"{}), : could not find function "{}\%>\%"{}}}

\begin{CodeChunk}
\begin{CodeOutput}
##    user  system elapsed 
##   0.342   0.005   0.348
\end{CodeOutput}
\begin{CodeOutput}
## [1] -47955
\end{CodeOutput}
\begin{CodeOutput}
## [1] 1015808
\end{CodeOutput}
\begin{CodeOutput}
## [1] 410825
\end{CodeOutput}
\begin{CodeOutput}
## [1] 0.404
\end{CodeOutput} 
\end{CodeChunk}

@@

\begin{CodeChunk}
\begin{CodeOutput}
##    user  system elapsed 
##  14.346   0.012  14.351
\end{CodeOutput}
\begin{CodeOutput}
## [1] 0.404
\end{CodeOutput} 
\end{CodeChunk}

@@

{\ttfamily\noindent\bfseries\color{errorcolor}{\#\# Error in knitr::kable(week, format = "{}latex"{}, align = c("{}rrrrrrrrrrrr"{}), : could not find function "{}\%>\%"{}}}

\begin{CodeChunk}
\begin{CodeOutput}
##    user  system elapsed 
##   1.941   0.009   1.949
\end{CodeOutput}
\begin{CodeOutput}
## [1] -54990
\end{CodeOutput}
\begin{CodeOutput}
## [1] 1e+07
\end{CodeOutput}
\begin{CodeOutput}
## [1] 1193
\end{CodeOutput}
\begin{CodeOutput}
## [1] 0.000119
\end{CodeOutput} 
\end{CodeChunk}

@@

\begin{CodeChunk}
\begin{CodeOutput}
##    user  system elapsed 
##  89.759   0.096  89.810
\end{CodeOutput}
\begin{CodeOutput}
## [1] 0.000117
\end{CodeOutput} 
\end{CodeChunk}

\subsection{A summary of the results}
We summarized the comparison results in the table \ref{tab:summary} and plot the test statistics in Figure %\ref{fig4}.
\begin{table}

@@

\begin{CodeChunk}
\begin{CodeInput}
R> data("swissRain", package = "geostatsp")
R> myRaster = geostatsp::squareRaster(swissBorder, 80)
R> myRaster
\end{CodeInput}
\begin{CodeOutput}
## class      : RasterLayer 
## dimensions : 51, 80, 4080  (nrow, ncol, ncell)
## resolution : 4358, 4358  (x, y)
## extent     : 2485351, 2833996, 1075013, 1297275  (xmin, xmax, ymin, ymax)
## crs        : +proj=somerc +lat_0=46.9524055555556 +lon_0=7.43958333333333 +k_0=1 +x_0=2600000 +y_0=1200000 +ellps=GRS80 +units=m +no_defs
\end{CodeOutput} 
\end{CodeChunk}

@@

Step 3, compute Mat\'ern covariance matrices using \fct{gpuBatchMatrix::maternBatch}, the returned Mat\'ern covariance matrices %$\Sigma_1, \Sigma_2, \dots, \Sigma_5$ correspond to the 5 parameter sets respectively, the matrices 
are each of size $4800 \times 4800$ and are stacked by row in the output \code{maternCov}. 
\begin{CodeChunk}
\begin{CodeOutput}
## [1] 16320  4080
\end{CodeOutput} 
\end{CodeChunk}
